---
output: github_document
---
# Betadiag

Diagnostics and Remedy Tools for Beta Diversity Analysis

Betadiag is an R package designed to diagnose, evaluate, and correct issues in beta diversity analysis, particularly when working with dissimilarity matrices from ecological and microbiome studies. It provides tools to assess matrix properties, perform ordination and statistical evaluation, and apply remedies to ensure valid analysis.

## Installation

To install the development version of Betadiag from GitHub:


```{r, message = FALSE}
# Install devtools if not already installed
# install.packages("devtools")
library(devtools)
devtools::install_github("bioscinema/Betadiag")
library(Betadiag)
```

## Key Features

- Check whether dissimilarity matrices are metric and Euclidean
- Assess triangle inequality violations, collinearity, and nonlinearity
- Convert dissimilarity matrices into Gram matrices
- Perform Principal Coordinate Analysis (PCoA) and eigenvalue decomposition
- Conduct PERMANOVA and pseudo R² analysis for association testing
- Apply Higham or Tikhonov corrections to enforce positive semi-definiteness
- Compute weighted UniFrac distances directly from phyloseq objects

## End-to-end demo: weighted UniFrac on IBD 16S data


```{r, warning=FALSE}
# Install Betadiag (dev) and dependencies ---------------------------------
# install.packages("devtools")
library(devtools)
devtools::install_github("bioscinema/Betadiag")

library(Betadiag)
library(phyloseq)
library(ggplot2)

# Load dataset ------------------------------------------------------------
load("RealData/IBD_16s_data_V4.RData")

# ---- Data cleaning ------------------------------------------------------
sample_data <- sample_data(phy1)
rows_with_na <- apply(
  sample_data[, c(1, 18, 50, 57, 115, 126, 127)],
  1,
  function(x) any(x %in% c("not providednot provided",
                           "-88",
                           "not provided",
                           "cd (from uc 7/17/2018)"))
)
physeq <- prune_samples(!rows_with_na, phy1)

metadata <- data.frame(
  diagnosis         = as.factor(physeq@sam_data$diagnosis),
  age_at_diagnosis  = as.numeric(physeq@sam_data$age_at__diagnosis),
  host_age          = as.numeric(physeq@sam_data$host_age),
  host_height       = as.numeric(physeq@sam_data$host_height),
  race              = physeq@sam_data$race,
  sex               = as.numeric(ifelse(physeq@sam_data$sex == "male", 1, 0)),
  smoking           = as.numeric(ifelse(physeq@sam_data$smoking == "n", 0, 1))
)

# ---- Diagnostics --------------------------------------------------------
wu_dist   <- phyloseq::distance(physeq, method = "wunifrac")
wu_matrix <- as.matrix(wu_dist)

wu.check <- check_distance(wu_matrix)
wu.check$is.metric
wu.check$is.Euclidean
wu.check$FNI
```


```{r}
plot(wu.check$collinearity.score,xlab = "Sample",ylab = "Collinearity Score") 
```

Collinearity plot – Every point sits at ≈ −1, meaning each sample’s distance vector is almost perfectly anti-parallel to the average vector. In practice, this indicates very little pairwise collinearity, so angle-preserving (metric) ordinations shouldn’t be distorted by redundant information.

```{r}
plot(wu.check$nonlinearity.score,xlab = "Sample",ylab = "Nonlinearity Score") 
```

Non-linearity plot – Scores vary widely, with several pronounced spikes, showing substantial non-linear distortion across samples. Linear methods such as classical PCoA may mis-represent true relationships; non-linear embeddings (t-SNE, UMAP) or a PSD remedy are advisable before interpretation.

```{r}
# ---- Baseline evaluation ------------------------------------------------
wu.evalution <- evaluate_beta(
  wu_matrix,
  as.data.frame(metadata[, 1]),  # Y
  as.data.frame(metadata[, -1]), # Z
  metadata
)
wu.evalution$PCo1.rate
wu.evalution$PCo2.rate
wu.evalution$PCoA.valid
wu.evalution$pseudo_F
wu.evalution$pseudo_R2
wu.evalution$permanova_p
wu.evalution$MiRKAT_R2
```



```{r}
# ---- Remedies -----------------------------------------------------------
# Higham projection
wu.higham           <- remedy_gram(wu_matrix, method = "Higham")
wu.higham.evalution <- evaluate_beta_gram(
  wu.higham,
  as.data.frame(metadata[, 1]),
  as.data.frame(metadata[, -1]),
  metadata
)
wu.higham.evalution$PCo1.rate
wu.higham.evalution$PCo2.rate
wu.higham.evalution$PCoA.valid
wu.higham.evalution$pseudo_F
wu.higham.evalution$pseudo_R2
wu.higham.evalution$permanova_p
wu.higham.evalution$MiRKAT_R2
```

```{r}
wu.higham.euclidean <- euclidean_check(wu.higham)
wu.higham.euclidean
```

```{r}
# Tikhonov ridge (ε = 0)
wu.tik           <- remedy_gram(wu_matrix, method = "Tikhonov", epsilon = 0)
wu.tik.evalution <- evaluate_beta_gram(
  wu.tik,
  as.data.frame(metadata[, 1]),
  as.data.frame(metadata[, -1]),
  metadata
)
wu.tik.evalution$PCo1.rate
wu.tik.evalution$PCo2.rate
wu.tik.evalution$PCoA.valid
wu.tik.evalution$pseudo_F
wu.tik.evalution$pseudo_R2
wu.tik.evalution$permanova_p
wu.tik.evalution$MiRKAT_R2
```

```{r}
wu.tik.euclidean <- euclidean_check(wu.tik)
wu.tik.euclidean
```


```{r}
# ---- Quick comparison table --------------------------------------------
data.frame(
  Matrix      = c("Raw", "Higham", "Tikhonov"),
  Euclidean   = c(wu.check$is.Euclidean,
                  wu.higham.euclidean$is.Euclidean,
                  wu.tik.euclidean$is.Euclidean),
  Pseudo_R2   = c(wu.evalution$pseudo_R2,
                  wu.higham.evalution$pseudo_R2,
                  wu.tik.evalution$pseudo_R2),
  PERM_pval   = c(wu.evalution$permanova_p,
                  wu.higham.evalution$permanova_p,
                  wu.tik.evalution$permanova_p)
)
```

